Reference Tracking Strategies of Deaf Adult Signers in Turkish Sign Language

### 2023-01-31 revised_manuscript.tex
updated

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Reference Tracking Strategies of Native and Late Signers in Turkish Sign Language},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

% Pandoc header
\usepackage{floatrow}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{float}
\floatplacement{figure}{H}
\floatplacement{table}{H}
\floatsetup[figure]{capposition=top}
\floatsetup[table]{capposition=top}
\usepackage{times}
\usepackage{smartdiagram}
\usepackage{ragged2e}
\usepackage{tikz}
\usepackage{xcolor}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{./sty/langsci-gb4e}
\usepackage{caption}
\usepackage{multicol}
\setlength{\columnsep}{-3cm}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

  \title{Reference Tracking Strategies of Native and Late Signers in
Turkish Sign Language}

  \begin{abstract}
  We investigated the reference tracking strategies among native and
  late adult deaf signers of Turkish Sign Language. Consistent with the
  theories of saliency and referential accessibility, regardless of
  their acquisiton groups, signers mainly used nominals and extension
  classifiers for introducing referents. To maintain a referent in the
  previous clause, signers chose null marking (e.g., constructed action,
  agreement and plain verbs). For topic shifts or re-introduced
  contexts, nominals and pronouns were chiefly favored although we
  observed very little pronominal use. As for age of acquisition
  effects, we only report limited over-redundancy for late signers who
  used null marking less compared to native signers especially for
  introduced and maintained contexts. Given our limited support for
  redundancy by late signers, we conclude that native signers might
  utilize the spatial affordances of the visual-spatial modality better
  than late signers only to a certain extent.
  \end{abstract}
   \begin{keyword} Referential accessibility, signed narratives,
pragmatic competence, delayed language acquisition, Turkish Sign
Language\end{keyword}
 \end{frontmatter}

@@

Narratives are extensive discourses that construct an important part of
human communication. In order to create a systematic and coherent
organization of experiences, narrators must choose appropriate
linguistic forms to refer to entities (Ariel, 1990; Givon, 1983; Gundel
et al., 1993): Certain forms such as nominal expressions are commonly
used for the first mention of a referent in a discourse, and other forms
such as pronouns and null marking may track previously mentioned
entities which could either be maintained across multiple clauses or
re-introduced back into the discourse. The selection of these referring
expressions (henceforth REs) is contingent on cognitive abilities, such
that individuals must rely on the activation of the working memory to
keep in mind previously referred expressions and long strings of
information (Bamberg, 1997; Morgan, 2005). Pragmatic abilities too play
a pivotal role in the tracking of referents. The form with the least
amount of redundant information must be selected based on the
accessibility levels of the referent (i.e., how active a referent is in
the addressee's mind given a point in discourse) (Ahn, 2019; Ariel,
1990). Because of these cognitive and pragmatic necessities, it is not
surprising to observe that producing a coherent narrative is one of the
latest milestones in language acquisition among children (Hickmann et
al., 1996). As can be expected, there is a plethora of research
examining how narrative competence is achieved in both monolingual and
bilingual acquisition contexts among children and adults. Although
narrative competence and referent selection have been studied
extensively for typically developing speakers (Aksu-Koç and
Nicolopoulou, 2015; Hickmann et al., 1996; Williams, 1988), deaf
individuals who acquired a sign language in atypical learning contexts
(i.e., deaf signers with linguistic deprivation) have remained mostly
understudied with an exception of handful of studies (Becker, 2009;
Cormier et al., 2013; Gür, 2018). Nevertheless, it is paramount that we
investigate and describe the effects of language deprivation among deaf
adults on narrative competence more in depth. This will help us
understand whether the loss of systematic and frequent language input in
the early years of life affect how deaf signers produce a narrative in
their adulthood. Given this, the present study will focus on deaf adult
signers with language deprivation and examine their reference tracking
strategies in signed narratives.

Sign Languages are natural languages of the Deaf communities all around
the world. They are independent languages, which do not derive from
majority spoken language(s) but emerge naturally when deaf people form a
community. Turkish Sign Language (Türk İşaret Dili -- TİD) is and has
been the officially recognized language of the Deaf community in Turkey
since 2005 (see Nuhbalaoglu, 2018 for a brief review of the historical
development of the language). Deaf or hearing children acquiring a SL
within a family where there is one or more deaf caregivers display the
same developmental milestones as hearing children acquiring a spoken
language in a hearing family (Chamberlain et al., 1999; Meier and
Newport, 1990). However, reports indicate that over 90\% of all deaf
individuals are born into hearing families who do not know how to sign
(i.e., those with language deprivation) (Woll, 2013). This means that
the auditory modality used in the household is not accessible to many
deaf children. That is why the age of acquisition of a first functional
language varies greatly among deaf children, with many of whom being
deprived of their first languages typically until after age 7 when they
start formal education in a deaf school. Despite this, many of the early
intervention programs for late signing children prioritize spoken
language development and discourage the use of a sign language
(Kemaloğlu and Kemaloğlu, 2012). Ministry of National Education (2020)
identifies 72 deaf schools in Turkey for the K-12 level. Even when deaf
children with hearing parents start their formal education in one of
these schools, they do not receive proper instruction in TİD since the
teachers themselves are mainly hearing and only have a very rudimentary
knowledge of a sign language (Ilkbasaran, 2015). First language
deprivation or delay has been found to have negative effects on certain
cognitive and linguistic measures among late signing children. To name a
few, delayed exposure to a first functional language is reported to
negatively affect deaf children and adults' morphosyntactic production
and processing (Boudreault and Mayberry, 2006; Cheng and Mayberry, 2021;
Cormier et al., 2012; Karadöller et al., 2017; Kayabaşı et al., 2020;
Newport, 1990) and narrative abilities (e.g., Becker, 2009). In our
overall research, we are concerned with the effects of this late
exposure in the ultimate adult forms of TİD users.

@@

Natural languages can be systematically used to keep track of previous
referents in the discourse and mark the changing cognitive statuses of
the REs (i.e., from being active to less active in the addressee's mind)
to create a more coherent narrative. Among the theoretical frameworks
that aim to explain that different linguistic forms are selected in
accordance with the addressee's mental representations of the referents
in the discourse is the Givenness Hierarchy (Gundel et al., 1993).
According to this model, given or familiar information is expressed in
more attenuated forms such as pronominal use and null marking since the
mental status of the expressed referent is already at the center of
attention, as indicated by the present contextual information (Chafe,
1976). The Givenness Hierarchy is akin to other previously proposed
models of reference control within the topicality and saliency paradigm
such as the Accessibility Hierarchy (Ariel, 1990), Topic Continuity
(Givon, 1983), and Familiarity Scale (Prince, 1981). Gundel et al.
(1993) argued that the distribution of forms is not random but abides by
the Gricean Maxim of Quantity (Grice, 1975). That is to say, individuals
must make the most informative and succinct choice possible to retain
economy of form. When multiple expressions are available for the same
referent as in an utterance like (\ref{ex:1}), phonetically simplest
form is selected with the least redundant information based on an
economy principle. In the utterance, REs for ``Eddie'' are given in
bold. All three REs (nominal, pronominal, and null marking) refer to the
same entity; however, the quantity of marking decreases as the
referent's accessibility increases. On this ground, for highly
accessible referents (i.e., those that are highly active in the mind),
linguistic forms with the least phonetic content are selected for the
least amount of processing effort.

@@

It is also possible to categorize REs into three according to their
discourse status (Gullberg, 2006). A referent can be introduced or
mentioned for the first time, maintained, that is, continued across two
or more clauses, or re-introduced into the discourse following a topic
shift. As a result, interlocutors mark some REs differently than others
based on the principles of reference accessibility. For spoken
languages, this observation is well established (Ahrenholz, 2005; Ariel,
1990; Azar and Özyürek, 2015; Chafe, 1976; Debreslioska et al., 2013;
Givon, 1983; Gundel et al., 1993; Hickmann et al., 1996). These research
indicate that the introduction and re-introduction of a referent are
often accomplished by full noun phrases which can be commonly modified
by an adjective or demonstrative. Maintenance, on the other hand, is
typically marked with pronominals and null markers since the continued
referent is still highly accessible to the addressee.

However, this particular observation is not peculiar to spoken
languages. There is an emerging body of literature that reports the
distribution of REs in signed narratives (Ahlgren and Bergman, 1994;
Czubek, 2017; Frederiksen and Mayberry, 2016; Hodge et al., 2019;
Morgan, 2005; Perniss and Özyürek, 2015; Swabey, 2002). Building on
Gundel et al. (1993)'s work, Swabey (2002) demonstrated that, like other
spoken languages such as Russian, Chinese, and Japanese, American Sign
Language (ASL) signers use bare nouns for introduced and re-introduced
discourse contexts whereas null marking is used for referents which are
``in-focus'' or very accessible to the addressee. Overt REs like
pronouns or nouns are disprefered for such referents. Swabey (2002)
concluded that ``native'' signers in her study abided by the Gricean
maxim of quantity and were thus economic in their reference tracking.
Similar findings have been reported for German Sign Language (Deutsche
Gebärdensprache - DGS) (Perniss and Özyürek, 2015) and British Sign
Language (BSL) (Morgan, 2005). It is important to note, however, that
these research either collapsed native and late signers into a single
group and labeled them as ``native-like'' or explicitly compared native
deaf signers with late deaf signers (with language deprivation) in terms
of narrative competence.

@@

The linguistic inventory for discourse referents can include nominal
constructions (bare nouns, fingerspelled nouns, and modified nouns),
null markers (plain verb, agreement verb, and constructed action),
classifier predicates (whole entity classifiers, body part classifiers,
handling classifiers, and extension classifiers), and overt pronouns
(pronominal index or IX) (Frederiksen and Mayberry, 2016). This interim
section describes such constructions attested for sign languages under
four main headings (nominals, pronouns, classifiers, and null marking).

@@

In TİD and many other sign languages, it is possible to use bare noun
phrases as in (\ref{ex:6a}) as well as fingerspelled nouns as in
(\ref{ex:6c}). Fingerspelling refers to the use of the manual alphabet
which is derived or borrowed from a surrounding written language (Kubuş,
2008). Nouns can also be modified by classifiers as in (\ref{ex:6b})
(Frederiksen and Mayberry, 2016) or by a prenominal or postnominal IX
functioning as a determiner or a demonstrative as in (\ref{ex:6d})
(Maclaughlin, 1997).

\hypertarget{pronouns}{%
\subsubsection{Pronouns}\label{pronouns}}

The IX sign (extended index finger) has various functions in sign
languages, one of which is the pronominal use which refers to
established nominals by pointing to a previously assigned abstract or
physical locus in the signing space when re-introducing a topic in
discourse (Emmorey, 1996). The same IX function has also been identified
for TİD (Sevinç, 2006). The example in (\ref{ex:7}) illustrates a
spatially-anchored pronominal reference for topic re-introduction. These
signs are thus rarely used for the first mention of an entity (Perniss
and Özyürek, 2015). Other interpretations consider pointing signs
without noun phrases as examples of demonstratives and not personal
pronouns (Koulidobrova and Lillo-Martin, 2016). It is also important to
mention here that in some spoken languages (e.g., German and Dutch),
demonstratives can also refer to persons (Bosch and Umbach, 2007) and
behave similarly to personal pronouns. However, for the purposes of the
present study, we will refer to IX constructions as pronominal IX and
particularly examine their non-deictic or personal pronoun realizations
as a possible discourse strategy to mainly mark topic shift among native
and late signers of TİD.

@@

Classifier predicates are highly complex morphological constructions
which are very common in sign languages (Supalla, 1986). These are
manually signed iconic representations of animate or inanimate objects,
and they embody several properties pertaining to the nominal referent
including its location in the signing space, movement, path, size, and
shape (Talmy, 2003). As in most sign languages hitherto studied,
classifier constructions are also commonly found in TİD (see Dikyuva et
al., 2015 for a review). Following the classification of
Engberg-Pedersen (1993), we distinguish between four types of classifier
constructions. Whole entity classifiers (WECL) represent the entirety of
any referent whereas body part classifiers (BPCL) only reflect a
specific part of an animate entity (e.g., limbs, head, tongue, teeth,
and so forth). Handling classifiers (HCL\footnote{As discussed before,
  an HCL is transitive in nature and depicts the handling action
  performed by ananimate agent. This is why most research tend to
  consider HCLs as an extension to constructed action (Cormier et al.,
  2013; Dudis, 2004).}) occur with transitive verbs and depict how an
object is handled or manipulated by an agent. Extension classifiers
(ExtCL), on the other hand, denote the size and shape of an entity by
following a trajectory to trace its shape. We exemplify these
constructions in (\ref{ex:8}).

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\bibliographystyle{elsarticle-harv}
\usepackage{graphicx}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Reference Tracking Strategies of Deaf Adult Signers in Turkish Sign Language},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls

@@

% Pandoc header
\usepackage{floatrow}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{float}
\floatplacement{figure}{H}
\floatplacement{table}{H}
\floatsetup[figure]{capposition=top}
\floatsetup[table]{capposition=top}
\usepackage{times}
\usepackage{smartdiagram}
\usepackage{ragged2e}
\usepackage{tikz}
\usepackage{xcolor}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{./sty/langsci-gb4e}
\usepackage{caption}
\usepackage{multicol}
\setlength{\columnsep}{-2cm}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

@@

  \title{Reference Tracking Strategies of Deaf Adult Signers in Turkish
Sign Language}

  \begin{abstract}
  We investigated the reference tracking strategies among deaf adults
  with signing deaf parents (DoD) and adult deaf signers with
  non-signing hearing parents or caregivers (DoH) who had late exposure
  to Turkish Sign Language (TİD). Consistent with the theories of
  saliency and referential accessibility, regardless of their acquisiton
  groups or parental hearing status, signers mainly used nominals and
  extension classifiers for introducing referents. To maintain a
  referent from the previous clause, the signers used zero anaphora
  (e.g., constructed action, agreement and plain verbs). For topic
  shifts or re-introduced contexts, nominals and pronouns were chiefly
  favored although we observed very little pronominal use. As for the
  effect of native acquisition from deaf caregivers, we only report
  limited over-redundancy for DoH signers who used zero anaphora less
  compared to DoD especially for introduced and maintained contexts. We
  conclude that DoH signers are still able to achieve native-like
  competency in terms of reference tracking in simple narratives but DoD
  signers utilize the spatial affordances of the visual-spatial modality
  better than DoH signers only to a certain extent.
  \end{abstract}
   \begin{keyword} Referential accessibility, signed narratives,
pragmatic competence, delayed language acquisition, Turkish Sign
Language\end{keyword}
 \end{frontmatter}

@@

Narratives are extensive discourses that form an important part of human
communication. In order to create a systematic and coherent organization
of experiences, narrators must choose appropriate linguistic forms to
refer to entities (Ariel, 1990; Givon, 1983; Gundel et al., 1993).
Certain forms such as nominal expressions are commonly used for the
first mention of a referent in a discourse, and other forms such as
pronouns and zero anaphora may track previously mentioned entities which
could either be maintained across multiple clauses or re-introduced back
into the discourse. The selection of these referring expressions
(henceforth REs) is contingent on cognitive abilities, such that
individuals must rely on the activation of the working memory to keep in
mind previously referred expressions and long strings of information
(Bamberg, 1997; Morgan, 2005). Pragmatic abilities too play a pivotal
role in the tracking of referents. The form with the least amount of
redundant information must be selected based on the accessibility levels
of the referent (i.e., how active a referent is in the addressee's mind
given a point in discourse) (Ahn, 2019; Ariel, 1990). Because of these
cognitive and pragmatic necessities, it is not surprising to observe
that producing a coherent narrative is one of the latest milestones in
language acquisition among children (Hickmann et al., 1996). As can be
expected, there is a plethora of research examining how narrative
competence is achieved in both monolingual and bilingual acquisition
contexts among children and adults. Although narrative competence and
referent selection have been studied extensively for typically
developing speakers (Aksu-Koç and Nicolopoulou, 2015; Hickmann et al.,
1996; Williams, 1988), deaf individuals who acquired a sign language in
atypical learning contexts (i.e., deaf signers with delayed exposure to
a sign language) have remained mostly understudied with an exception of
a handful of studies (Becker, 2009; Cormier et al., 2013; Gür, 2018).
Nevertheless, it is paramount that we investigate and describe the
effects of late exposure to a sign language among deaf adults growing up
in a signing vs.~non-signing environment on narrative competence more in
depth. This will help us understand whether the lack of systematic and
frequent language exposure in the early years of life affect how deaf
signers produce a narrative in their adulthood. Given this, the present
study will focus on two groups of signers and examine their reference
tracking strategies in signed narratives: (i) deaf of deaf (DoD) adult
signers (i.e., signers who have deaf caregivers and natively acquired a
sign language from birth) and (ii) deaf of hearing (DoH) adult signers
(i.e., signers who have hearing non-signing caregivers and state to have
started acquiring a sign language later in life).

Sign languages are the natural languages of the Deaf communities all
around the world. They are independent languages, which do not derive
from majority spoken language(s) but emerge naturally when deaf people
form a community. Turkish Sign Language (Türk İşaret Dili -- TİD) is and
has been the officially recognized language of the Deaf community in
Turkey since 2005 (see Nuhbalaoglu, 2018 for a brief review of the
historical development of the language). Deaf or hearing children
acquiring a SL within a family where there is one or more deaf
caregivers display the same developmental milestones as hearing children
acquiring a spoken language in a hearing family (Chamberlain et al.,
1999; Meier and Newport, 1990). However, reports indicate that over 90\%
of all deaf individuals are born into hearing families (DoH) who do not
know how to sign (Woll, 2013). This means that the auditory modality
used in the household is not accessible to many deaf children. That is
why the age of acquisition of a first functional language varies greatly
among deaf children, with many of whom being deprived of their first
languages typically until after age 7 when they start formal education
in a deaf school, especially in the context of Turkey (Sari, 2005).
Despite this, many of the early intervention programs for late signing
children prioritize spoken language development and discourage the use
of a sign language (Kemaloğlu and Kemaloğlu, 2012). The Ministry of
National Education (2020) identifies 72 deaf schools in Turkey for the
K-12 level. Even when deaf children with hearing parents start their
formal education in one of these schools, they do not receive proper
instruction in TİD since the teachers themselves are mainly hearing and
only have a very rudimentary knowledge of a sign language (Ilkbasaran,
2015; Sari, 2005). In some cases, it has been argued that late exposure
to a sign language among certain DoH signers has negative effects on
some cognitive and linguistic measures. To name a few, delayed exposure
to a first functional language is reported to negatively affect deaf
children and adults' morphosyntactic production and processing
(Boudreault and Mayberry, 2006; Cheng and Mayberry, 2021; Cormier et
al., 2012; Karadöller et al., 2017; Kayabaşı and Gökgöz, 2022; Newport,
1990) and narrative abilities (e.g., Becker, 2009). In our overall
research, we are concerned with the effects of this late exposure in the
ultimate adult forms of TİD users.

@@

Natural languages can be systematically used to keep track of previous
referents in the discourse and mark the changing cognitive statuses of
the REs (i.e., from being active to less active in the addressee's mind)
to create a more coherent narrative. Among the theoretical frameworks
that aim to explain that different linguistic forms are selected in
accordance with the addressee's mental representations of the referents
in the discourse is the Givenness Hierarchy (Gundel et al., 1993).
According to this model, given or familiar information is expressed in
more attenuated forms such as pronouns and zero anaphora (i.e., there is
no explicit mention of the referent) since the mental status of the
expressed referent is already at the center of attention, as indicated
by the present contextual information (Chafe, 1976). The Givenness
Hierarchy is akin to other previously proposed models of reference
control within the topicality and saliency paradigm such as the
Accessibility Hierarchy (Ariel, 1990), Topic Continuity (Givon, 1983),
and Familiarity Scale (Prince, 1981). Gundel et al. (1993) argued that
the distribution of forms is not random but abides by the Gricean Maxim
of Quantity (Grice, 1975). That is to say, individuals must make the
most informative and succinct choice possible to retain economy of form.
When multiple expressions are available for the same referent as in an
utterance like (\ref{ex:1}), phonetically simplest form is selected with
the least redundant information based on an economy principle. In the
utterance, REs for ``Eddie'' are given in bold. All three REs (nominal,
pronominal, and zero anaphora) refer to the same entity; however, the
quantity of marking decreases as the referent's accessibility increases.
On this ground, for highly accessible referents (i.e., those that are
highly active in the mind), linguistic forms with the least phonetic
content are selected for the least amount of processing effort.

@@

It is also possible to categorize REs into three according to their
discourse status (Gullberg, 2006). A referent can be introduced or
mentioned for the first time, maintained, that is, continued across two
or more clauses, or re-introduced into the discourse following a topic
shift. As a result, interlocutors mark some REs differently than others
based on the principles of reference accessibility. For spoken
languages, this observation is well established (Ahrenholz, 2005; Ariel,
1990; Azar and Özyürek, 2015; Chafe, 1976; Debreslioska et al., 2013;
Givon, 1983; Gundel et al., 1993; Hickmann et al., 1996). These research
indicate that the introduction and re-introduction of a referent are
often accomplished by full noun phrases which can be commonly modified
by an adjective or demonstrative. Maintenance, on the other hand, is
typically marked with pronominals and zero anaphora since the continued
referent is still highly accessible to the addressee.

However, this particular observation is not peculiar to spoken
languages. There is an emerging body of literature that reports the
distribution of REs in signed narratives (Ahlgren and Bergman, 1994;
Czubek, 2017; Frederiksen and Mayberry, 2016; Hodge et al., 2019;
Morgan, 2005; Perniss and Özyürek, 2015; Swabey, 2002). Building on the
work by Gundel et al. (1993), Swabey (2002) demonstrated that, like
other languages such as Russian, Chinese, and Japanese, American Sign
Language (ASL) signers use bare nouns for introduced and re-introduced
discourse contexts whereas zero anaphora is used for referents which are
``in-focus'' or very accessible to the addressee. Overt REs like
pronouns or nouns are disprefered for such referents. Swabey (2002)
concluded that ``native'' signers in her study abided by the Gricean
maxim of quantity and thus used more economic forms of reference
tracking. Similar findings have been reported for German Sign Language
(Deutsche Gebärdensprache - DGS) (Perniss and Özyürek, 2015) and British
Sign Language (BSL) (Morgan, 2005). It is important to note, however,
that this research either collapsed DoD and DoH signers into a single
group and labeled them as ``native-like'' or explicitly compared native
deaf signers with late deaf signers (with delayed exposure to a sign
language) in terms of narrative competence.

@@

The linguistic inventory for discourse referents can include nominal
constructions (bare nouns, fingerspelled nouns, and modified nouns),
null or zero
anaphors\footnote{Zero anaphora does not refer to the lack of a referent. It only alludes to the non-presence of a more explicit anaphoric form for the referent such as a nominal, pronominal, or classifiers.}
(predicated by a plain verb, an agreement verb, or constructed action),
classifier predicates (whole entity classifiers, body part classifiers,
handling classifiers, and extension classifiers), and overt pronouns
(pronominal index or IX) (Frederiksen and Mayberry, 2016). This interim
section describes such forms attested for sign languages under four main
headings (nominals, pronouns, classifiers, and zero anaphora).

@@

In TİD and many other sign languages, it is possible to use bare nouns
as in \ref{ex:2a} as well as fingerspelled nouns as in \ref{ex:2b}.
Fingerspelling refers to the use of the manual aphabet to spell all or
part of the letters of a word borrowed from a surrounding written
language (Kubuş, 2008). Nouns can also be modified by a prenominal or
postnominal extended index finger pointing sign, glossed as IX,
functioning as a determiner or a demonstrative as in \ref{ex:2c}
(Maclaughlin, 1997) or by a modifying classifier within the same NP as
in \ref{ex:2d} (Frederiksen and Mayberry,
2016).\footnote{Spoken languages use pronouns, determines and demonstratives where sign languages use a pointing sign, usually glossed as IX, which assumes similar functions (Lillo-Martin and Klima, 1990). In addition, the pointing sign is used to referentially anchor a referent to a location in signing space.}

\newpage
\begin{exe}
\ex
\begin{xlist}
\ex \label{ex:2a} Full noun \\\glll
{} \includegraphics[width=35pt]{pictures/p1.png} {} \\ \textbf{Gloss:} \textsc{mouse\footnotemark} \textsc{sick} \\ 
\textbf{Status:} Intro. \\
\glt \textbf{Context:} Narration 6, Event 1. Here, the signer introduces the mouse, Jerry, at the beginning of the narration with a bare noun.
\glt \textbf{Translation:} There is a sick mouse.
\vspace{0.35cm}

\ex \label{ex:2b} Fingerspelled noun \\\glll
{} \includegraphics[width=40pt]{pictures/p2.1.png} \includegraphics[width=40pt]{pictures/p2.2.png} \includegraphics[width=40pt]{pictures/p2.3.png} {} \\ \textbf{Gloss:} \textsc{t-} \textsc{-o-} \textsc{-m} \textsc{come.back} \\ 
\textbf{Status:} Re-intro. \\
\glt \textbf{Context:} Narration 7, Event 2. The signer introduces Tom two sentences before. She introduces the black cat in the preceding clause and then here re-introduces Tom by using fingerspelling.
\glt \textbf{Translation:} Tom comes back.
\vspace{0.35cm}

\ex \label{ex:2c} Noun modified by index (IX) \\\glll
{} \includegraphics[width=45pt]{pictures/p3.1.png} \includegraphics[width=40pt]{pictures/p3.2.png} {} {} {} \\ \textbf{Gloss:} \textsc{ix} \textsc{cat} \textsc{pipe.extend} \textsc{go.through} \textsc{exit} \\ 
\textbf{Status:} Re-intro. \\
\glt \textbf{Context:} Narration 1, Event 6. The signer introduces the cat in the first clause of the first event and re-introduces it several times including this last event of the narrative.
\glt \textbf{Translation:} The cat goes through and exits the thin pipe.
\vspace{0.35cm}

\ex \label{ex:2d} Noun modified by classifier (CL) \\\glll
{} \includegraphics[width=35pt]{pictures/p4.1.png} \includegraphics[width=35pt]{pictures/p4.2.png} \\ \textbf{Gloss:} \textsc{soap} \textsc{soap.cl} \\ 
\textbf{Status:} Intro. \\
\glt \textbf{Context:} Narration 5, Event 3. Here the signer introduces the soap to the discourse. The signer produces the lexical sign, \textsc{soap}, plus a classifier, \textsc{soap.cl}.
\glt \textbf{Translation:} The soap, which is like this.
\end{xlist}
\end{exe}

\footnotetext{We use small capitals to indicate sign glosses.}

\hypertarget{referential-use-of-space-and-pronouns}{%
\subsubsection{Referential Use of Space and
Pronouns}\label{referential-use-of-space-and-pronouns}}

In addition to being a modifier of a noun, the IX sign can also be
pronominal by referring to established nominals through pointing to a
previously assigned abstract or physical locus in the signing space when
re-introducing a topic in discourse (Emmorey, 1996). This function has
also been identified for TİD (Sevinç, 2006). The example in \ref{ex:3}
illustrates this spatially-anchored pronominal use.

\ea 
\ea \label{ex:3} Pronominal \\\glll
{} \includegraphics[width=35pt]{pictures/p5.png} {} {} {} {} \\ \textbf{Gloss:} \textsc{ix} \textsc{look} \textsc{cat.sad(constructed action)} \textsc{sad} \textsc{sick} \\ 
\textbf{Status:} Re-intro. \\
\glt \textbf{Context:} Narration 9, Event 1: The signer introduces the mouse then signs an action of the cat. Here, he re-introduces the mouse with a pointing sign that functions as a pronoun.
\glt \textbf{Translation:} It (the mouse) looks sad and sick.
\z \z

These pronominal signs are rarely used for the first mention of an
entity (Perniss and Özyürek, 2015). Other researchers consider pointing
signs without noun phrases as examples of demonstratives and not
personal pronouns (Koulidobrova and Lillo-Martin, 2016). It is also
important to mention here that in some spoken languages (e.g., German
and Dutch), demonstratives can also refer to persons (Bosch and Umbach,
2007) and behave similarly to personal pronouns. However, for the
purposes of the present study, we will refer to IX signs as pronominal
IX and particularly examine their non-deictic or personal pronoun
realizations as a possible discourse strategy to mainly mark topic shift
(re-introduction) among DoD and DoH signers of TİD.

@@

In sign languages, a verbal classifier morpheme is morpho-phonologically
encoded by a handshape. This handshape is articulated co-temporally with
a movement that encodes the verbal meaning (Benedicto and Brentari,
2004; Supalla, 1990). Semantically, the classifiers denote sets of
objects based on criteria such as animacy, size, shape, body-part,
texture and sometimes function (Talmy, 2003). The classifier articulated
simultaneously with a movement is called a classifier predicate and the
sentence in which such a predicate is used is usually called a
classifier construction in sign language research. As in most sign
languages hitherto studied, classifier constructions are also commonly
found in TİD (see Dikyuva et al., 2015 for a review). Following the
classification of (Engberg-Pedersen, 1993), in TİD, we distinguish
between four types of classifier handshapes. Whole entity classifiers
(WECL) represent the entirety of a referent (e.g.~an entire flat
horizontal entity, or an upright entity) (e.g., \ref{ex:4a}) whereas
body part classifiers (BPCL) represent a specific part of an animate
entity (e.g., limbs, head, mouth, tongue, or teeth, etc.)
(e.g.,\ref{ex:4b}). Handling classifiers (HCL) occur with transitive
verbal roots and represent both the size, shape and surface of an object
and how that object is handled or manipulated by an agent (e.g.,
\ref{ex:4c}). Extension classifiers (ExtCL), on the other hand,
represent the size and shape of an entity by following a trajectory to
trace its shape (e.g., \ref{ex:4d}). We exemplify these handshapes with
the discourse status of the referent they track and a description of the
previous discourse when available.

\begin{exe}
\ex
\begin{xlist}
\ex \label{ex:4a} Whole entity classifier  (WECL) \\\glll
{} \includegraphics[width=40pt]{pictures/p6.png} \\ \textbf{Gloss:} \textsc{two.mice.go.together} \\ 
\textbf{Status:} Maintain. \\
\glt \textbf{Context:} Narration 4, Event 2: The signer introduces Jerry and the robot mouse in the preceding clause. She maintains reference to them by using a whole entity classifier on each hand. 
\glt \textbf{Translation:} (Jerry and the robot mouse from the previous context) go together.
\vspace{0.35cm}

\ex \label{ex:4b} Body part classifier (BPCL) \\\glll
{} \includegraphics[width=40pt]{pictures/p7.png} \\ \textbf{Gloss:} \textsc{cat.move.thin.legs} \\ 
\textbf{Status:} Maintain. \\
\glt \textbf{Context:} Narration 1, Event 6: The signer re-introduces the cat three clauses before and maintains reference to him in the two clauses before and here with a body part classifier that refers to the legs of the cat.
\glt \textbf{Translation:} (The cat) walks (i.e. moving his thin legs forward).
\vspace{0.35cm}

\ex \label{ex:4c} Handling classifier (HCL) \\\glll
{} {} {} {} {} \includegraphics[width=40pt]{pictures/p8.png} \\ \textbf{Gloss:} {} \textsc{rope see-3a} \textsc{soon} \textsc{go} \textsc{cat.pull.rope} \\ 
\textbf{Status:} Maintain. \\
\glt \textbf{Context:} Narration 5, Event 2: The signer introduces Tom three clauses before. She maintains reference to him in the preceding two clauses and here she still maintains reference to him by using a handling classifier on both hands.
\glt \textbf{Translation:} When (he/Tom) sees the rope, (he) soon goes to pull it.
\vspace{0.35cm}

\ex \label{ex:4d} Extension classifier (ExtCL) \\\glll
{} \includegraphics[width=80pt]{pictures/p9.png} \\ \textbf{Gloss:} \textsc{cat.extend} \\ 
\textbf{Status:} Maintain. \\
\glt \textbf{Context:} Narration 1, Event 6: The signer re-introduces Tom two clauses before and she maintains reference to him with constructed action in the previous clause and here with an extension classifier.
\glt \textbf{Translation:} (The cat) extends thin.
\end{xlist}
\end{exe}

\hypertarget{zero-anaphora}{%
\subsubsection{Zero Anaphora}\label{zero-anaphora}}
